Gusto: a toolkit for compatible finite element dynamical cores

### 2024-07-24 source/paper.tex
tiny bit more progress

We illustrate the use of FML for this example in three steps: 1)
define and label the residual; 2) build the explicit and implicit
solvers; 3) run the timestepping loop. Since the intention of this
example is to demonstrate FML rather than the compatible finite
element discretisation, we use piecewise quadratic continuous Lagrange
polynomials, $P_2$, for the finite element discretisation as this
greatly simplifies the forms. The finite element discretisation of
equation \ref{eq:1dadvdiff} is then
\begin{equation}\label{eq:weak-advdiff}
  \int_\Omega vq \diff x + c \int_\Omega v q_x \diff x + \kappa \int_\Omega v_x q_x \diff x = 0 \quad \forall v \in P_2,
\end{equation}
where $v\in P_2$ is the test function and the diffusion term has been
integrated by parts.

%\begin{align}\label{eq:weak-advdiff}
%  \int_\Omega vq^* \diff x &= \int_\Omega vq^n \diff x - \Delta t c \int_\Omega v q_x^n \diff x, &\quad \forall v \in P^2, \\
%  \int_\Omega vq^{n+1} \diff x  + \Delta t \kappa \int_\Omega v_x q_x^{n+1} \diff x &= \int_\Omega vq^* \diff x, &\quad \forall v \in P^2,
%\end{align}
%where $v$ is the test function and the diffusion term has been
%integrated by parts.

Listing \ref{lst:define-residual} presents the code to define this
spatial discretisation of the residual with the $P_2$ function space
denoted by \texttt{V}. Here we omit the code that sets up the mesh,
function spaces and parameters $c$ and $\kappa$, although this can be
found in the supplementary materials. The first two lines define the
test function \texttt{v} and function \texttt{q}, corresponding to $v$
and $q$ in equation \ref{eq:weak-advdiff}. The next four lines define
the residual, i.e. the left hand side of equation
\ref{eq:weak-advdiff}, using the FML labels \texttt{time\_derivative},
\texttt{transport} and \texttt{diffusion} to tag the corresponding
terms. The final line labels the function \texttt{q} as the subject of
this residual. Note that the labels used for tagging just take the UFL
form as an argument and their value is simply the default, \texttt{True}, in
contrast to the \texttt{subject} label that has the value \texttt{q}.

@@

To define the solvers for the explicit and implicit steps, we make use
of the predefined mappings \texttt{drop} and \texttt{keep}

>>>>

We illustrate the use of FML for this example in three steps: 1)
define and label the weak form of the residual; 2) build the solvers
for the explicit and implicit systems; 3) run the timestepping
loop. Since the intention of this example is to demonstrate FML rather
than the compatible finite element discretisation, we use space of piecewise
quadratic continuous Lagrange polynomials, denoted by $P_2$, for the finite
element discretisation as this greatly simplifies the forms. The
finite element discretisation of equation \ref{eq:1dadvdiff} is then
\begin{equation}\label{eq:weak-advdiff}
  \int_\Omega vq_t \diff x + c \int_\Omega v q_x \diff x + \kappa \int_\Omega v_x q_x \diff x = 0, \quad \forall v \in P_2,
\end{equation}
where $v\in P_2$ is the test function and the diffusion term has been
integrated by parts.

Listing \ref{lst:define-residual} presents the code to define this
spatial discretisation of the residual with the $P_2$ function space
denoted by \texttt{V}. Here we omit the code that sets up the mesh,
function spaces and parameters $c$ and $\kappa$, although this can be
found in the supplementary materials. The first two lines define the
test function \texttt{v} and function \texttt{q}, corresponding to $v$
and $q$ in equation \ref{eq:weak-advdiff}. The next four lines define
the residual, i.e. the left hand side of equation
\ref{eq:weak-advdiff}, using the FML labels \texttt{time\_derivative},
\texttt{transport} and \texttt{diffusion} to tag the corresponding
terms. The final line labels the function \texttt{q} as the subject of
this residual, i.e. the prognostic field. Note that the labels used
for tagging just take the UFL form as their argument and their value is
simply the default, \texttt{True}, in contrast to the \texttt{subject}
label that has the value \texttt{q}.

@@

To define the solvers for the explicit and implicit steps we need to create UFL forms for the left and right sides, given by the following equations
\begin{align}
  \label{eq:explicit-advdiff}
  \int_\Omega vq^* \diff x &= \int_\Omega vq^n \diff x - \Delta t c \int_\Omega v q_x^n \diff x, &\quad \forall v \in P^2, \\
  \label{eq:implicit-advdiff}
  \int_\Omega vq^{n+1} \diff x  + \Delta t \kappa \int_\Omega v_x q_x^{n+1} \diff x &= \int_\Omega vq^* \diff x, &\quad \forall v \in P^2.
\end{align}
To do this we use the \texttt{label\_map} method to loop over the
terms in the residual making use of the predefined mappings
\texttt{drop} and \texttt{keep} to either remove or retain the term
based on the value of the filter. This code is given in listings
\ref{lst:define-explicit} and \ref{lst:define-implicit}. In listing
\ref{lst:define-explicit} we create the left and right hand sides of
equation \ref{eq:explicit-advdiff}. The left hand side should only
contain the time derivative term so the filter checks if each term has
the label \texttt{time\_derivative}, keeping the one that does and
dropping all others. To make this the left hand side of a
\texttt{LinearVariationalProblem}, we replace the subject with a trial
function using the predefined \texttt{replace\_subject} function, a
lightweight wrapper around the UFL replace method which extracts the
subject of the term and replaces it with the new expression, in this
case simply a trial function \texttt{q\_trial}. \JScomment{Is the
  labelling for the rhs the clearest? maybe we should keep the terms
  that we want rather than assuming that dropping diffusion gives the
  right thing?}
